Orienting the causal relationship between imprecisely measured traits using genetic instruments

### 2017-03-11 docs/manuscript.Rmd
updated

read_chunk("~/repo/cit_measurement_error/scripts/mr_directionality_analysis.R")
opts_chunk$set(warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE)

@@

Observational measures of the human phenome are growing ever more abundant, but using these data to make causal inference is notoriously susceptible to many pitfalls, with basic regression-based techniques unable to distinguish a true causal association from reverse causation or confounding [@Phillips1991; @DaveySmith2002; @DaveySmith2004]. In response to this, genetic instrumentation has emerged as a technique for improving the reliability of causal inference in observational data, and with the coincident rise in genome-wide association studies it is now a prominent tool that is applied in several different guises [@DaveySmith2004; @Millstein2009; @Aten2008; @Waszak2015]. However, potential pitfalls remain and one that is often neglected is the influence of non-differential measurement error on the reliability of causal inference.

Measurement error is the difference between the measured value of a quantity and its true value. This study focuses specifically on non-differential measurement error where all strata of a measured variable have the same error rate, which can manifest as changes in scale or levels of imprecision (noise) of the measurement. Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009]. Array technology is now commonly used to obtain high throughput phenotyping at low cost, but comes with the problem of having imperfect resolution, for instance methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value due to imperfect sensitivity [@Harper2013; @Chen2013a]. Relatedly, if the measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].

@@

Mediation-based approaches employ genetic instruments (typically single nucleotide polymorphisms, SNPs) to orient the causal direction between the exposure and the outcome. If a SNP is associated with an exposure, and the exposure is associated with some outcome, then it logically follows that in this simple three-variable scenario the indirect influence of the SNP on the outcome will disappear when conditioning on the exposure. Here, the exposure 'mediates' the association between the SNP and the outcome, providing information about the causal influence of the exposure on the outcome. This forms the basis of a number of methods such as genetical genomics [@Schadt2005], the regression-based causal inference test (CIT) [@Millstein2009; @Millstein2016], a structural equation modelling (SEM) implementation in the NEO software [@Aten2008], and various other methods including Bayesian approaches [@Waszak2015]. They have been employed by a number of recent publications that make causal inferences in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015; @Yuan2014; @Tang2014; @Hong2015].

@@

Assuming biological knowledge of genetic associations can be problematic because if there exists a putative association between two variables, with the SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the primary effect of the SNP (i.e. for which of the two variables is the SNP a valid instrument? `r cfig("dag_figure")`). By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, such that if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure. Such a situation can arise in many scenarios. Genome-wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015]. If multiple instruments are available for an hypothesised exposure, which is increasingly typical for complex traits that are analysed in large GWAS consortia, then techniques can be applied to mitigate these issues [@DaveySmithHemani2014]. But these techniques cannot always be applied in the case of determining causal directions between 'omic measures where typically only one cis-acting SNP is known. For example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level (`r cfig("dag_figure")`)?

MR has some important advantages over the mediation-based approaches. The mediation-based approaches require that the exposure, outcome and instrumental variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013]. This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes, and dramatically expands the breadth of possible phenotypic relationships that can be evaluated [@Hemani2016]. It is important to note that if there are any confounders influencing both the exposure and outcome, the mediation-based approach of adjusting the outcome for the exposure to nullify the association between the SNP and the outcome will not work. This is because adjusting the outcome by the exposure induces a collider effect between the SNP and outcome [@Hernan2004], and the in order to fully abbrogate this association one must also adjust for all (hidden or otherwise) confounders. MR does not suffer from this problem because it does not test for association through adjustment. Additionally, when MR assumptions are satisfied the method is robust to there being measurement error in the exposure variable [@Pierce2012]. Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error in the exposure [@Ashenfelter1994], whereas it has been noted that both classic mediation-based analyses [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation-based methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.

@@

We can model a system whereby some exposure $x$ has a causal influence $\beta_x$ on an outcome $y$ such that

@@

The $\alpha_*$ terms represent intercepts, and the $\epsilon_*$ denote random error. Mediation-based analyses that test whether $x$ causally relates to $y$ impinge on evaluating if the influence of $g$ on $y$ can be accounted for by conditioning on $x$, such that

@@

where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$. MR analysis estimates the causal influence of $x$ on $y$ by using the instrument as a proxy for $x$, such that

@@

Measurement error of an exposure can be modeled as a transformation of the true value that leads to the observed value, $x_o = f(x)$. For example, following Pierce and VanderWeele [-@Pierce2012] we can define

@@

where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$ by altering its scale, and $\epsilon_{mx}$ represents the imprecision (or noise) in the measurement of $x$. The same model of measurement error can be applied to the outcome variable $y$. In this study we assume there is no measurement error in the SNP and measurement error in the exposure and the outcome are uncorrelated.

@@

The CIT method [@Millstein2009] is implemented in the R package *R/cit* [@Millstein2016]. The methodology of the CIT is as follows. Assume an exposure $x$ is instrumented by a SNP $g$, and the exposure $x$ causes an outcome $y$, as described above. The following tests are then performed:

@@

where $y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for $x$, where $x$ is assumed to mediate the association between the SNP and the outcome. The 4th condition is formulated as an equivalence testing problem that is estimated using simulations, comparing the estimate against from the data against empirically obtained estimates for simulated variables where the independence model is true (full details are given in [@Millstein2009]). We note here that this approach is liable to fail, even when there is a true causal relationship, when confounders are present due to inducing collider effects.

@@

In these analyses the *cit.cp* function was used to obtain an omnibus p-value. To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT for each of two tests - $p_{CIT, x \rightarrow y}$, was estimated for the direction of $x$ causing $y$ (Model 1), and for the direction of $y$ causing $x$, $p_{CIT, y \rightarrow x}$ (Model 2). The results from each of these methods can then be used in combination to infer the existance and direction of causality. For some significance threshold $\alpha$ there are four possible outcomes from these two tests, and their interpretations are as follows:

@@

Two stage least squares (2SLS) is a commonly used technique for performing MR when the exposure, outcome and instrument data are all available in the same sample. A p-value for this test, $p_{MR}$, was obtained using the R package $R/systemfit$ [@Henningsen2007].

@@

and a p-value, $p_{Steiger}$ is generated from the $Z$ value to indicate the probability of obtaining the observed difference in correlations $\rho_{gx}$ and $\rho_{gy}$ under the null hypothesis that both correlations are identical.

The existence of causality and its direction is inferred based on combining information from the MR analysis and the Steiger test. The MR analysis indicates whether there is a potential causal relationship ($p_{MR}$), and the Steiger test indicates the direction ($sign(Z)$) of the causal relationship and the confidence of the direction ($p_{Steiger}). For the purposes of compiling simulation results, these can be combined using an arbitrary $\alpha = 0.05$ value:

- If $p_{Steiger} < \alpha$ and $p_{MR} < \alpha$ and $Z > 0$ then a causal association for the correct model is accepted
- If $p_{Steiger} < \alpha$ and $p_{MR} < \alpha$ and $Z < 0$ then a causal association for the incorrect model is accepted
- Otherwise no evidence for a causal relationship

@@

The Steiger test for inferring if $x \rightarrow y$ is based on evaluating $\rho_{gx} > \rho_{gy}$. However, $\rho_{gx}$ and $\rho_{gy}$ are underestimated if $x$ and $y$ are measured imprecisely. If, for example, $x$ has lower measurement precision than $y$ then we might empirically obtain $\rho_{g,x_o} < \rho_{g,y_o}$ because $\rho_{g,x_o}$ could be underestimated more than $\rho_{g,y_o}$.

As we show in Appendix 2 it is possible to infer the bounds of measurement error on $x_o$ or $y_o$ given known genetic associations. The maximum measurement error of $x_o$ is $\rho_{g,x_o}$, because it is known that at least that much of the variance has been explained in $x_o$ by $g$. The minimum is 0, denoting perfectly measured trait values (the same logic applies to $y_o$). It is possible to simulate what the inferred causal direction would be for all values within these bounds.

@@

## Two sample MR

@@

All analysis was performed using the R programming language [@RCoreTeam2015] and code is made available at [github url to go here]() and implemented in the MR-Base (http://wwww.mrbase.org) platform [@Hemani2016].

@@

Thus an observational study will find $cov(g, y_o - \hat{y_o}) = 0$ when the true model is causal only when $D = 1$. Therefore, if there is any measurement error that incurs imprecision (i.e. $var(\epsilon_{mx}) \neq 0$) then there will remain an association between $g$ and $y_o | x$, which is in violation of the the 4th condition of the CIT. Note that scale transformation of $x$ or $y$ without any incurred imprecision is insufficient to lead to a violation of the test statistic assumptions, and henceforth mention of measurement error will relate to imprecision unless otherwise stated.

@@

We performed simulations to compare the performance of MR against CIT in detecting a causal association between simulated variables under different levels of imprecision simulated in the exposure. `r cfig("causality_exists_tpr")` shows the true positive rates between the CIT and MR for detecting a causal association. We observe that the CIT has lower power in all cases, with performance declining as measurement imprecision increases in the exposure. When MR assumptions are satisfied, notably that it is known on which of $x$ and $y$ the SNP $g$ has a primary influence, the performance of MR in detecting an association is unrelated to measurement error in the exposure. Measurement error in the outcome does reduce power, but does not induce a substantive difference in performance between CIT and MR.

@@

We used simulations to explore the performance of the MR Steiger approach in comparison to CIT in terms of the rate at which a strong causal relationship is obtained for the correct direction of causality, and the rate at which a strong causal relationship is obtained where the reported direction of causality is incorrect. Simulations were performed for two models, one for a "causal model" where there was a causal relationship between $x$ and $y$; and one for a "non-causal model" where $x$ and $y$ were not causally related, but had a confounded association induced by the SNP $g$ influencing a confounder variable $u$.

`r cfig("cit_mr_comparison_figure")`a shows that, for the "causal model", when $d < 0$ the MR analysis is indeed liable to infer the wrong direction of causality, and that this erroneous result is more likely to occur with increasing sample size. However, the CIT is in general more fallable to reporting a robust causal association for the wrong direction of causality. When $d > 0$ is satisfied we observe that in most cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality. The CIT, unlike the Steiger test for MR, is able to distinguish the "non-causal model" from the "causal model" (Methods, `r cfig("cit_mr_comparison_figure")`b), but it is evident that measurement error will often lead the CIT to evaluate erroneously this confounding model as being causal.

@@

We used the Steiger test to infer the direction of causality between `r nrow(qtl_orig)` putative associations between DNA methylation levels and gene expression levels and found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a), but assuming no or equal measurement error, DNA methylation levels were the predominant causal factor ($p = `r signif(shakhtest4$p.value, 2)`$). The median reliability (R) of the `r nrow(qtl_orig)` tests was `r round(shakh_median_reliability, 2)` (95% CI `r round(shakh_r_loci, 2)` - `r round(shakh_r_upci, 2)`). We then went on to predict the causal directions of the associations for varying levels of systematic measurement error for the different platforms. `r cfig("shakhplot")`a shows that the qualitative result of which of DNA methylation or gene expression is more commonly the causal factor is very strongly determined by measurement error.

We performed two sample MR [@Pierce2013] for each association in the direction of causality inferred by the Stieger test. We observed that the sign of the MR estimate was generally in the same direction as the Pearson correlation coefficient reported by Shakhbazov et al [-@Shakhbazov2016]  (`r cfig("shakhplot")`b). There was a moderate correlation between the absolute magnitudes of the causal correlation and the observational Pearson correlation (r = `r round(cor(abs(shakhbazov$mr_r[real_index]), abs(shakhbazov$pearson[real_index])), 2)`). Together these inferences suggest that even for relatively low level associations it is important to use causal inference methods over observational associations to infer causal effect sizes.

@@

Researchers are often confronted with the problem of making causal inferences using a statistical framework on observational data. In the epidemiological literature issues of measurement error in mediation analysis is relatively well explored [@Cole2014]. Our analysis extends this to related methods such as CIT that are widely used in predominantly 'omic data. These methods are indeed liable to the same issues as standard mediation based analysis, and specifically we show that as measurement error in the exposure variable increases, CIT is likely to have reduced statistical power, and liable to infer the wrong direction of causality. We also demonstrate that, though unintuitive, increasing sample size does not resolve the issue, rather it leads to more extreme p-values for the model that predicts the wrong direction of causality.

>>>>

read_chunk("../scripts/mr_directionality_analysis.R")
opts_chunk$set(warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE)

@@

Observational measures of the human phenome are growing ever more abundant, but using these data to make causal inference is notoriously susceptible to many pitfalls, with basic regression-based techniques unable to distinguish a true causal association from reverse causation or confounding [@Phillips1991; @DaveySmith2002; @DaveySmith2004]. In response to this, the use of genetic associations to instrument traits has emerged as a technique for improving the reliability of causal inference in observational data, and with the coincident rise in genome-wide association studies it is now a prominent tool that is applied in several different guises [@DaveySmith2004; @Millstein2009; @Aten2008; @Waszak2015]. However, potential pitfalls remain and one that is often neglected is the influence of non-differential measurement error on the reliability of causal inference.

Measurement error is the difference between the measured value of a quantity and its true value. This study focuses specifically on non-differential measurement error where all strata of a measured variable have the same error rate, which can manifest as changes in scale or measurement imprecision (noise). Such variability can arise through a whole plethora of mechanisms, which are often specific to the study design and difficult to avoid [@Houle2011; @Hernan2009]. Array technology is now commonly used to obtain high throughput phenotyping at low cost, but comes with the problem of having imperfect resolution, for instance methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value due to imperfect sensitivity [@Harper2013; @Chen2013a]. Relatedly, if the measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].

@@

Mediation-based approaches employ genetic instruments (typically single nucleotide polymorphisms, SNPs) to orient the causal direction between the exposure and the outcome. If a SNP is associated with an exposure, and the exposure is associated with some outcome, then it logically follows that in this simple three-variable scenario the estimated direct influence of the SNP on the outcome will be zero when conditioning on the exposure. Here, the exposure completely mediates the association between the SNP and the outcome, providing information about the causal influence of the exposure on the outcome. This forms the basis of a number of methods such as genetical genomics [@Schadt2005], the regression-based causal inference test (CIT) [@Millstein2009; @Millstein2016], a structural equation modelling (SEM) implementation in the NEO software [@Aten2008], and various other methods including Bayesian approaches [@Waszak2015]. They have been employed by a number of recent publications that make causal inferences in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015; @Yuan2014; @Tang2014; @Hong2015].

@@

Assuming biological knowledge of genetic associations can be problematic because if there exists a putative association between two variables, with the SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the primary effect of the SNP (i.e. for which of the two variables is the SNP a valid instrument? `r cfig("dag_figure")`). By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, such that if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure. Genome-wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015]. If multiple instruments are available for an hypothesised exposure, which is increasingly typical for complex traits that are analysed in large GWAS consortia, then techniques can be applied to mitigate these issues [@DaveySmithHemani2014]. But these techniques cannot always be applied in the case of determining causal directions between 'omic measures where typically only one cis-acting SNP is known. For example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level (`r cfig("dag_figure")`)?

MR has some important advantages over the mediation-based approaches. First, the mediation-based approaches require that the exposure, outcome and instrumental variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013]. This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes, and dramatically expands the breadth of possible phenotypic relationships that can be evaluated [@Hemani2016]. Second, the mediation-based approach of adjusting the outcome for the exposure to nullify the association between the SNP and the outcome is affected by unmeasured confounding of the exposure and outcome. This is because adjusting the outcome by the exposure induces a collider effect between the SNP and outcome [@Hernan2004], and the in order to fully abrogate this association one must also adjust for all (hidden or otherwise) confounders. MR does not suffer from this problem because it does not test for association through adjustment. Third, when MR assumptions are satisfied the method is robust to there being measurement error in the exposure variable [@Pierce2012]. Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error in the exposure [@Ashenfelter1994], whereas it has been noted that both classic mediation-based analyses [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation-based methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.

@@

We model a system whereby some exposure $x$ has a causal influence $\beta_x$ on an outcome $y$ such that

@@

The $\alpha_*$ terms represent intercepts, and henceforth can be ignored. The $\epsilon_*$ terms denote random error, assumed independently and normally distributed with mean zero. Mediation-based analyses that test whether $x$ causally relates to $y$ rely on evaluating whether the influence of $g$ on $y$ can be accounted for by conditioning on $x$, such that

@@

where $\hat{y} = \hat{\beta}_x x$. MR analysis estimates the causal influence of $x$ on $y$ by using the instrument as a proxy for $x$, such that

@@

Measurement error of an exposure can be modeled as a transformation of the true value ($x$) that leads to the observed value, $x_o = f(x)$. For example, following Pierce and VanderWeele [-@Pierce2012] we can define

@@

where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$ by altering its scale, and $\epsilon_{mx}$ represents the imprecision (or noise) in the measurement of $x$. The same model of measurement error can be applied to the outcome variable $y$. In this study we assume there is no measurement error in the SNP, and that measurement error in the exposure and the outcome are uncorrelated.

@@

First we describe how the CIT method [@Millstein2009] is implemented in the R package *R/cit* [@Millstein2016]. The methodology of the CIT is as follows. Assume an exposure $x$ is instrumented by a SNP $g$, and the exposure $x$ causes an outcome $y$, as described above. The following tests are then performed:

@@

where $y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for $x$, where $x$ is assumed to mediate the association between the SNP and the outcome. The 4th condition is formulated as an equivalence testing problem that is estimated using simulations, comparing the estimate against from the data against empirically obtained estimates for simulated variables where the independence model is true (full details are given in [@Millstein2009]). We note here that this approach is liable to fail, even when there is a true causal relationship, when confounders of the exposure and outcome are present, as these will induce collider bias.

@@

Now we describe how we used the CIT method in our simulations. The *cit.cp* function was used to obtain an omnibus p-value. To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT for each of two tests - $p_{CIT, x \rightarrow y}$, was estimated for the direction of $x$ causing $y$ (Model 1), and for the direction of $y$ causing $x$, $p_{CIT, y \rightarrow x}$ (Model 2). The results from each of these methods can then be used in combination to infer the existance and direction of causality. For some significance threshold $\alpha$ there are four possible outcomes from these two tests, and their interpretations are as follows:

@@

Two stage least squares (2SLS) is a commonly used technique for performing MR when the exposure, outcome and instrument data are all available in the same sample. A p-value for this test, $p_{MR}$, was obtained using the $systemfit$ function in the R package $R/systemfit$ [@Henningsen2007]. Note that the value of $p_{MR}$ is identical when using the same genetic variant to instrument the influence of the exposure $x$ on the outcome $y$, or erroneously, instrumenting the outcome $y$ on the exposure $x$.

@@

and a p-value, $p_{Steiger}$ is generated from the $Z$ value to indicate the probability of obtaining a difference between correlations $\rho_{gx}$ and $\rho_{gy}$ at least as large as the one observed, under the null hypothesis that both correlations are identical.

The existence of causality and its direction is inferred based on combining information from the MR analysis and the Steiger test. The MR analysis indicates whether there is a potential causal relationship ($p_{MR}$), and the Steiger test indicates the direction ($sign(Z)$) of the causal relationship and the confidence of the direction ($p_{Steiger}$). For the purposes of compiling simulation results, these can be combined using an arbitrary $\alpha = 0.05$ value:

- If $p_{Steiger} < \alpha$ and $p_{MR} < \alpha$ and $Z > 0$ then a causal association for the correct model is accepted, $x \rightarrow y$
- If $p_{Steiger} < \alpha$ and $p_{MR} < \alpha$ and $Z < 0$ then a causal association for the incorrect model is accepted, $y \rightarrow x$
- Otherwise if $p_{Steiger} > \alpha$ or $p_{MR} > \alpha$, neither model is accepted

@@

The Steiger test for inferring if $x \rightarrow y$ is based on evaluating $\rho_{gx} > \rho_{gy}$. However, $\rho_{gx}$ (or $\rho_{gy}$) are underestimated if $x$ (or $y$) are measured imprecisely. If, for example, $x$ has lower measurement precision than $y$ then we might empirically obtain $\rho_{g,x_o} < \rho_{g,y_o}$ because $\rho_{g,x_o}$ could be underestimated more than $\rho_{g,y_o}$.

As we show in Appendix 2 it is possible to infer the bounds of measurement error on $x_o$ or $y_o$ given known genetic associations. The maximum measurement imprecision of $x_o$ is $\rho_{g,x_o}$, because it is known that at least that much of the variance has been explained in $x_o$ by $g$. The minimum is 0, denoting perfectly measured trait values (the same logic applies to $y_o$). It is possible to simulate what the inferred causal direction would be for all values within these bounds.

@@

## Applied example using two sample MR

@@

All analysis was performed using the R programming language [@RCoreTeam2015] and code is made available at [https://github.com/explodecomputer/causal-directions](https://github.com/explodecomputer/causal-directions) and implemented in the MR-Base (http://wwww.mrbase.org) platform [@Hemani2016].

@@

Thus an observational study will find $cov(g, y_o - \hat{y_o}) = 0$ when the true model is causal only when $D = 1$. Therefore, if there is any measurement error that incurs imprecision in $x$ (i.e. $var(\epsilon_{mx}) \neq 0$) then there will remain an association between $g$ and $y_o | x_o$, which is in violation of the the 4th condition of the CIT. Note that scale transformation of $x$ or $y$ without any incurred imprecision is insufficient to lead to a violation of the test statistic assumptions, and henceforth mention of measurement error will relate to imprecision unless otherwise stated.

@@

We also performed simulations to compare the performance of MR against CIT in detecting a causal association between simulated variables under different levels of imprecision simulated in the exposure. `r cfig("causality_exists_tpr")` shows the true positive rates between the CIT and MR for detecting a causal association. We observe that the CIT has lower power in all cases, with performance declining as measurement imprecision increases in the exposure. When MR assumptions are satisfied, notably that it is known on which of $x$ and $y$ the SNP $g$ has a primary influence, the performance of MR in detecting an association is unrelated to measurement error in the exposure. Measurement error in the outcome does reduce power, but does not induce a substantive difference in performance between CIT and MR.

@@

We used simulations to explore the performance of the MR Steiger approach in comparison to CIT in terms of the rate at which evidence of a causal relationship is obtained for the correct direction of causality, and the rate at which evidence of a causal relationship is obtained where the reported direction of causality is incorrect. Simulations were performed for two models, one for a "causal model" where there was a causal relationship between $x$ and $y$; and one for a "non-causal model" where $x$ and $y$ were not causally related, but had a confounded association induced by the SNP $g$ influencing a confounder variable $u$.

`r cfig("cit_mr_comparison_figure")`a shows that, for the "causal model", the MR analysis is indeed liable to infer the wrong direction of causality when $d < 0$, and that this erroneous result is more likely to occur with increasing sample size. However, the CIT is in general more fallable to reporting a robust causal association for the wrong direction of causality. When $d > 0$ we find that in most cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality. The CIT, unlike the Steiger test for MR, is able to distinguish the "non-causal model" from the "causal model" (Methods, `r cfig("cit_mr_comparison_figure")`b), but it is evident that measurement error will often lead the CIT to identify the causal model as true, when in fact the underlying model is this non-causal model.

@@

We used the Steiger test to infer the direction of causality between DNA methylation and gene expression levels between `r nrow(qtl_orig)` putative associations. We found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a), but assuming no or equal measurement error, DNA methylation levels were the predominant causal factor ($p = `r signif(shakhtest4$p.value, 2)`$). The median reliability (R) of the `r nrow(qtl_orig)` tests was `r round(shakh_median_reliability, 2)` (5%-95% quantiles `r round(shakh_r_loci, 2)` - `r round(shakh_r_upci, 2)`). We then went on to predict the causal directions of the associations for varying levels of systematic measurement error for the different platforms. `r cfig("shakhplot")`a shows that the conclusions about the direction of causality between DNA methylation and gene expression are very sensitive to measurement error.

We performed two sample MR [@Pierce2013] for each association in the direction of causality inferred by the Stieger test. We observed that the sign of the MR estimate was generally in the same direction as the Pearson correlation coefficient reported by Shakhbazov et al [-@Shakhbazov2016]  (`r cfig("shakhplot")`b). There was a moderate correlation between the absolute magnitudes of the causal correlation and the observational Pearson correlation (r = `r round(cor(abs(shakhbazov$mr_r[real_index]), abs(shakhbazov$pearson[real_index])), 2)`). Together these inferences suggest that even in estimating associations between 'omic' variables, which are considered to be low level phenotypes, it is important to use causal inference methods over observational associations to infer causal effect sizes.

@@

Researchers are often confronted with the problem of making causal inferences using a statistical framework on observational data. In the epidemiological literature issues of measurement error in mediation analysis are relatively well explored [@Cole2014]. Our analysis extends this to related methods such as CIT that are widely used in predominantly 'omic data. These methods are indeed susceptible to the same problem as standard mediation based analysis, and specifically we show that as measurement error in the (true) exposure variable increases, CIT is likely to have reduced statistical power, and liable to infer the wrong direction of causality. We also demonstrate that, though unintuitive, increasing sample size does not resolve the issue, rather it leads to more extreme p-values for the model that predicts the wrong direction of causality.
